\name: A Non-Parametric, Hierarchical \\ Wavelength-Calibration Model for a Precision Spectrograph

### 2020-05-14 paper/paper.tex
A bit of clean-up

The second simple idea follows from the observation that contemporary \eprv\ instruments are incredibly stable.  Temperature-controlled, fiber-fed spectrographs vary only slightly over the night or season.  \lz{This suggests that such stabilized spectrographs vary along only} a small number of axes in what you might call ``calibration space'', or the (very high dimensional) space of all possible wavelength solutions.  That is, not only are the spectrographs stable, but they also have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph was located for each exposure.

@@

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven model for the wavelength solution.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically determined variations in the stabilized spectrograph.  \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases, but we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \mb{The motivating ideas behind this project are true for almost every astronomical spectrograph. (Is this true?)}

@@

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} to be a 2D, 9\textsuperscript{th}-order polynomial where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here the y-dependence is dropped as that dependence is carried by spectral order $m$.  \lz{The line position information is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.}  The coefficients $c_{nij}$ are interpolated to time $t_n$ of exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

@@

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that they can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$,
and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  When this calibration structure is used to deliver a wavelength solution, we will have to invert $x(\lambda,m,n)$ into $\lambda(x,m,n)$.  \mb{(Explain why this inversion is necessary?)}

@@

\subsection{Dimensionality Reduction: De-Noising of Calibration Frames} \label{sec:denoising}
Within a given range of stability, \name\ will use calibration images to  1) determine the space in which the instrument varies and 2) where in the accessible calibration space the spectrograph existed for each exposure.  For each calibration exposure, $n$, \name\ requires a full list of lines, $(\lambda,m)$ that are expected in each calibration exposure.  Each line is uniquely defined by a combination of echelle order, $m$, and ``true'' or theoretical wavelength, $\lambda$.  \lz{There are many strategies for identifying calibration line positions and matching them to their assigned wavelengths; this problem is left out of scope for this work.}

@@

In the implementation of \name\ presented here, the magnitudes of the principal components are interpolated with respect to time.  New magnitudes $a_n'k$ are found for the midpoint time of non-calibration exposures and used to construct the calibration state of the spectrograph for that exposure.  Using interpolated magnitudes, $a_{n;k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, \name\ can construct a set of calibration lines, $x(\lambda,m,n') $, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.  \lz{The best interpolation scheme for \expres\ is discussed below in Section \ref{sec:choice_avt}, but may vary for different instruments.}

@@

 \lz{(Mention degeneracy with prescribed line center and wavelength solution here?  Or in discussion?)}

@@

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{Figures/all_results.pdf}
\caption{\lz{(change to be b/w safe)} Difference in predicted and assigned wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The standard deviation of the residuals for each test is given in the top-right corner in each method's corresponding color.}
\label{fig:testHists}
\end{figure}

@@

\name\ levies the incredible stability of contemporary EPRV instruments and high density of lines made available by new calibration sources, such as LFCs and etalons, to achieve more accurate wavelengths.  Denser calibration lines allow us to move to more flexible, interpolated wavelength solutions.  Stabilized spectrograph hardware with few degrees of freedom allow us to use all calibration images in a given generation of stability to constrain the accessible calibration space of the spectrograph.

We have described only one, fairly simplistic implementation of \name\ here.  There are many options for both the de-noising and interpolation steps.  \lz{(Maybe something to discuss at a future meeting; I'm not actually entirely clear on other possibilities)}.  For instance, a Gaussian process model could be constructed to determine wavelengths in place of a spline. \lz{(?)}  \lz{Something about going Bayesian?}

\mb{(Is PCA a ``choice'' or an ``extension'')}
An advantage of this implementation, where PCA is applied to line positions from all LFC exposures, is the ability to isolate exposures that exhibit errant variation, which is typically associated with flawed exposures.  This allowed us to quickly vet for problematic LFC exposures, which otherwise would have required visual inspection of all 1200+ LFC exposures.  In a classic framework where each calibration exposure is treated independently, these aberrant exposures would likely have persisted undetected and could potentially sway the resultant wavelength solutions.

>>>>

The second simple idea follows from the observation that contemporary \eprv\ instruments are incredibly stable.  Temperature-controlled, fiber-fed spectrographs vary only slightly over the night or season.  This suggests that such stabilized spectrographs vary along only a small number of axes in what you might call ``calibration space'', or the (very high dimensional) space of all possible wavelength solutions.  That is, not only are the spectrographs stable, but they also have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph was located for each exposure.

@@

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven model for the wavelength solution.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically determined variations in the stabilized spectrograph.  \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind \eprv\ instruments and \eprv\ science cases, but we expect \name\ to have applications for other kinds of spectrographs in other contexts.  \name\ should be applicable to almost every astronomical spectrograph.

@@

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} to be a 2D, 9\textsuperscript{th}-order polynomial where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
Here the y-dependence is dropped as that dependence is carried by spectral order $m$.  The line position information is therefore uniquely identified by echelle order $m$ and pixel in the dispersion direction, $x$.  The coefficients $c_{nij}$ are interpolated to time $t_n$ of exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains its $c_{nij}$ coefficients independently.

@@

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that they can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$,
and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  When this calibration structure is used to deliver a wavelength solution, we will have to invert $x(\lambda,m,n)$ into $\lambda(x,m,n)$.

@@

\subsection{Dimensionality Reduction: De-Noising of Calibration Frames} \label{sec:denoising}
Within a given range of stability, \name\ will use calibration images to  1) determine the space in which the instrument varies and 2) where in the accessible calibration space the spectrograph existed for each exposure.  For each calibration exposure, $n$, \name\ requires a full list of lines, $(\lambda,m)$ that are expected in each calibration exposure.  Each line is uniquely defined by a combination of echelle order, $m$, and ``true'' or theoretical wavelength, $\lambda$.  There are many strategies for identifying calibration line positions and matching them to their assigned wavelengths; this problem is left out of scope for this work.

@@

In the implementation of \name\ presented here, the magnitudes of the principal components are interpolated with respect to time.  New magnitudes $a_n'k$ are found for the midpoint time of non-calibration exposures and used to construct the calibration state of the spectrograph for that exposure.  Using interpolated magnitudes, $a_{n;k}$, and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, \name\ can construct a set of calibration lines, $x(\lambda,m,n') $, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.  The best interpolation scheme for \expres\ is discussed below in Section \ref{sec:choice_avt}, but may vary for different instruments.

@@

@@

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and assigned wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The standard deviation of the residuals for each test is given in the top-right corner in each method's corresponding color.}
\label{fig:testHists}
\end{figure}

@@

\name\ levies the incredible stability of contemporary EPRV instruments and high density of lines made available by new calibration sources, such as LFCs and etalons, to achieve more accurate wavelengths.  Denser calibration lines allow us to move to more flexible, interpolated wavelength solutions.  Stabilized spectrograph hardware with few degrees of freedom allow us to use all calibration images in a given generation of stability to constrain the accessible calibration space of the spectrograph.  We have described only one, fairly simplistic implementation of \name\ here.  There are many options for both the de-noising and interpolation steps, as mentioned in Section \ref{sec:choices}.

An advantage of this implementation, where PCA is applied to line positions from all LFC exposures, is the ability to isolate exposures that exhibit errant variation, which is typically associated with flawed exposures.  This allowed us to quickly vet for problematic LFC exposures, which otherwise would have required visual inspection of all 1200+ LFC exposures.  In a classic framework where each calibration exposure is treated independently, these aberrant exposures would likely have persisted undetected and could potentially sway the resultant wavelength solutions.
